The Numerical Accuracy of Computing Electrostatic Interaction in the Inhomogeneous and Correlated Molecular Systems: for the Ewald Summation, SPME and Staggered Mesh Ewald Methods.

### 2012-07-11 longRange.hetero/alltext/publish/resubmit/lett.tex
finish first revise, add figures

\begin{document}

@@

?

\textit{
In the introduction, the authors should probably take the perspective that SPME is a special case of P3M.  In the opinions of the developers of SPME, the derivation differs from P3M although it leads to similar results, but even TA Darden acknowledges that the reciprocal space component of the non-smooth PME algorithm was inspired by P3M.  It is probably best to phrase SPME as such and not offend anyone.  The interlacing technique was first applied in the 1970s, as the authors have cited.  It was originally a means of reducing the memory requirements of these calculations—electrostatics on the two grids were solved serially.  Finally, the sentence “We prove that the error is decomposed…” should be changed to “We show that the error may be approximately decomposed…” to indicate that the estimator is, in the end, approximate but relevant to real systems.
}\\

not done\\

@@

We delete Fig.3 and plot the charge distribution of Example 2 together
with Example 1 in the new Fig.1. Changes to the caption and
descriptions are maded correspondingly.\\

\textit{
The authors are reasonable to focus on the error arising from the
reciprocal space approximations in the various P3M-related methods;
the error arising from the direct space part of the calculation is
simpler to assess.  However, because the authors give so much
attention to SPME and the Staggered Mesh method which, as published,
built on SPME, the authors should include some plot of the effect of
grid spacing, Gaussian smoothing function width (Ewald coefficient),
and interpolation order -- these are the determinants of the accuracy of
the reciprocal space calculation in these fast methods.  How does the
estimator do in terms of getting the various trends correct?  What is
the effect of correlated charge positions on the error and the
estimates under changing interpolation order?  This is at the heart of
the difference between ik- and analytic differentiation that the
authors have otherwise carefully investigated.  It would also be
helpful to see a plot of the error estimator’s performance as a
function of the “near neighbor” definition.  Define near neighbors
strictly by length, and plot the error estimator's accuracy as a
continuous function of the near neighbor cutoff.  Extra credit for
detailing the computation time as a function of the same parameter!
}\\

@@

Page 7, line 23:
The definition of a discrete set of nearest neighbors $\Omega_j$ might not readily avaible in all cases. For example, for simple liquids it may be more feasible to use the radial distribution function $g(r)$ up to a certain radius $r_c$, where $g(r) \approx 1$ for $r > c_c$. Then the sum has to be replaced by an integral over $r$.  Because of the close relation of $\rho^{(2)}$ and $g(r)$, the equations should be easy to adapt to such a notation.

Page 15, line 45:
I believe I saw that the for interlaced P$^3$M algorithm, which is very similar to the staggered SMPE algorithm,  the approximation to leading order of the error estimate is not justified, at least for higher precisions. This approximation could lead to a visible deviation from the true errors. In this context it might be useful to recalculate the numerical examples at higher precision. Usually, deviations start to be visible from about an RMS of $10^{-4}$.

Page 16, line 34:
Again, In the very similarly structured interlaced P$^3$M algorithm,  there are differences in accuracy between the $i\mathbf{k}$- and analytical differentiation at higher precisions.At this point it might not be justified  to assume that both can be treated in the same way. Similarly to the approximation of the $i\mathbf{k}$ error formula, it would be useful to do calculations at higher precision than those presented in the paper, and compare them

The manuscript contains a few misprints, and the language could be polished on a few occasions by a native speaker.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{correlation.eps}
  \caption{The schematic plot of two interacting water molecules.
    When we calculate the force error of the left oxygen atom, the
    correlation with its bonded hydrogens are not considered, but
    the correlations within any neighboring molecule are considered,
    as indicated by the solid lines with two arrows.}
  \label{fig:1}
\end{figure}

We define the force error by inserting a testing charge into the
considered system, and comparing the calculated and exact force on
it. In the considered water system (Example 3), the subtle difference
between this setting to the realistic case is that the testing charge
has no correlation with the charges in the system, while any real
charge is bonded to two neighboring charges (via H-O bond and H-O-H
angle).  Therefore, the error estimate does not consider the
correlation between a charge and the its bonded neighbors, but it does
consider all bonding correlations within any neighboring molecule (see
Fig.~\ref{fig:1} of this letter). This is consistent with the usually
setting of bonded pair atoms, as mentioned by the reviewer, the
electrostatic interaction between them should not be calculated.

We therefore added the following sentence to make this point clearer:
\textbf{
  We define the force error by inserting a
testing charge into the water system, and comparing the
calculated and exact force on it. This definition implicitly assumes
the independency between the testing particle and the
system. Therefore, it sould be noted that the correlation error
estimate (20)
neglects the possible bonding
of the testing particle to any hydrogen/oxygen atom, and
calculates all bonded correlations
within its {neighboring} molecules. This is consistent
with the usual setting of the bonded atoms: the electrostatic interaction
between them is considered to be \emph{not} calculated.}

@@

As far as we concerned, it is not possible to test this point
directly, but we think it is a likely explanation to the observed
phenomenon.  Consider the extreme case: wihin one molecule, the
hydorgen and oxygen atoms locate at the same point (strongest
correlation), then the positive and negative charge cancels, so the
error vanishes. In case of TIP3P water model, the H-O distence is 0.1~nm and
the H-H distance is 0.16~nm, which are much smaller than the typical
center-of-mass distance between two neighboring molecules (0.31~nm).
We added this point to the paper.
\\

>>>>

\begin{document}
\noindent
Dear Prof. Rothlisberger,\\

Thank you very much for the reports. We would like to thank also the
reviewers for the positive reports and for the suggestions, which help
us improving the quality of the manuscript. The revisions are
reported below in detail.\\

In the revised manuscript, all the additions are marked in red, to help the
reviewers to located them immediately in the text.\\

@@

We have modified the introduction according to this point.
Please see the 4-th line of the second paragraph of the introduction:
\textbf{
  This problem can be solved by repeating and comparing
  the force calculation by
  using different parameter sets
  on a representative snapshot of the system,
  see eg. Ref.[16].
An other approach is the 
  in-depth theoretical study
of the error introduced by the algorithms, namely the
error estimate,
by which the error is described by a function of the working
parameters...
}\\

\textit{ In the introduction, the authors should probably take the
  perspective that SPME is a special case of P3M.  In the opinions of
  the developers of SPME, the derivation differs from P3M although it
  leads to similar results, but even TA Darden acknowledges that the
  reciprocal space component of the non-smooth PME algorithm was
  inspired by P3M.  It is probably best to phrase SPME as such and not
  offend anyone.  The interlacing technique was first applied in the
  1970s, as the authors have cited.  It was originally a means of
  reducing the memory requirements of these
  calculations—electrostatics on the two grids were solved serially.
  Finally, the sentence ``We prove that the error is decomposed...''
  should be changed to ``We show that the error may be approximately
  decomposed...'' to indicate that the estimator is, in the end,
  approximate but relevant to real systems.
}

We have added the comment (see the 12-th line of the 1st paragraph
of the introduction:
\textbf{It has been demonstrated that the SPME is actually
  a special case of the P3M method [6,7].}
The sentence ``We prove that the error is decomposed...'' has been
changed to \textbf{We show that the error may be approximately
  decomposed...}
\\

@@

We delete the old Fig.3,
and plot the charge distribution of Example 2 together
with Example 1 in the new Fig.2. Changes to the caption and
descriptions have been made correspondingly.\\

\textit{
The authors are reasonable to focus on the error arising from the
reciprocal space approximations in the various P3M-related methods;
the error arising from the direct space part of the calculation is
simpler to assess.  However, because the authors give so much
attention to SPME and the Staggered Mesh method which, as published,
built on SPME, the authors should include some plot of the effect of
grid spacing, Gaussian smoothing function width (Ewald coefficient),
and interpolation order -- these are the determinants of the accuracy of
the reciprocal space calculation in these fast methods.  How does the
estimator do in terms of getting the various trends correct?  What is
the effect of correlated charge positions on the error and the
estimates under changing interpolation order?  This is at the heart of
the difference between ik- and analytic differentiation that the
authors have otherwise carefully investigated.  It would also be
helpful to see a plot of the error estimator’s performance as a
function of the “near neighbor” definition.  Define near neighbors
strictly by length, and plot the error estimator's accuracy as a
continuous function of the near neighbor cutoff.  Extra credit for
detailing the computation time as a function of the same parameter!
}

We add new Fig. 9-12 to show how the error estimates
with bonded nearest neighbor approximations are able to catch the
trends of the real errors with various Ewald coefficient $\beta$, direct space
cut-off radius $r_c$, reciprocal space grid spacing $h$ and B-spline
interpolation order $n$.
% These figures show the error estimate
% with the bonded nearest neighbor approximation correctly catches the
% trend of changing working parameters.

In the revised manuscript,
we consider two types of the correlations:
(1) the bonded correlation
that originates from the rigid H-O
bond and H-O-H angle within one molecule,
and (2) the
non-bonded correlation or the
intermolecular correlation that originates from
a complex coupling of various intermolecular interaction.
Estimates for the bonded correlation
have already been developed by the old manuscript,
while the non-bonded estimate
is newly developed according to the reviewers' suggestion.
Please note that
both the bonded and non-bonded nearest neighbor approximations
require some uniformity of the correlation in the system:
the bonded estimate requires the uniformity of the water
direction preference, while the non-bonded estimate requires the uniformity
of the density distribution and the isotropicity of the system.
Otherwise the cost of the error estimate would be $\mathcal O(N^2\log N)$.
For the non-bonded correlation,  the nearest neighbor approximation 
is expressed in terms of the radial distribution
functions (RDFs). In example 3, a global RDF is not well defined,
so we calculated the RDF and nearest neighbor approximation
only for the liquid phase, see the dotted line in Fig.7. By adding
the non-bonded correlation, the quality of the error estimate
is further improved.
Also the convergence with respect to
the neighboring range $R_g$ is investigated (Fig.8). 
\\

@@

\textit{ Page 7, line 23: The definition of a discrete set of nearest
  neighbors $\Omega_j$ might not readily avaible in all cases. For
  example, for simple liquids it may be more feasible to use the
  radial distribution function $g(r)$ up to a certain radius $r_c$,
  where $g(r) \approx 1$ for $r > r_c$. Then the sum has to be
  replaced by an integral over $r$.  Because of the close relation of
  $\rho^{(2)}$ and $g(r)$, the equations should be easy to adapt to
  such a notation.  }

In the revised manuscript, the correlation error are considered
in two types: 
(1) the bonded correlation
that originates from the rigid H-O
bond and H-O-H angle within one molecule,
and (2) the
non-bonded correlation or the
intermolecular correlation that originates from
a complex coupling of various intermolecular interaction.
Estimates for the bonded correlation
have already been developed by the old manuscript,
while the non-bonded estimate
is newly developed according to the reviewers' suggestion.
Please note that
both the bonded and non-bonded nearest neighbor approximations
require some uniformity of the correlation in the system:
the bonded estimate requires the uniformity of the water
direction preference, while the non-bonded estimate requires the uniformity
of the density distribution and the isotropicity of the system, which
are also the prerequisites for reducing $\rho^{(2)}$ to $g(r)$.
%Otherwise the cost of the error estimate would be $\mathcal O(N^2\log N)$.
For the non-bonded correlation,  the nearest neighbor approximation 
is expressed in terms of the radial distribution
functions (RDFs) in the new Eqn. (21) and (22).
In example 3, a global RDF is not well defined
(because the system is inhomogeneous),
so we calculated the RDF and nearest neighbor approximation
only for the liquid phase, see the dotted line in Fig.7. By adding the
non-bonded correlation, the quality of the error estimate
is further improved.
Also the convergence with respect to
the neighboring range $R_g$ (the $r_c$ in the reviewer's comment)
is investigated (Fig.8). 
\\

\textit{ Page 15, line 45: I believe I saw that the for interlaced
  P$^3$M algorithm, which is very similar to the staggered SMPE
  algorithm, the approximation to leading order of the error estimate
  is not justified, at least for higher precisions. This approximation
  could lead to a visible deviation from the true errors. In this
  context it might be useful to recalculate the numerical examples at
  higher precision. Usually, deviations start to be visible from about
  an RMS of $10^{-4}$.  }

\textit{ Page 16, line 34: Again, In the very similarly structured
  interlaced P$^3$M algorithm, there are differences in accuracy
  between the $i\mathbf{k}$- and analytical differentiation at higher
  precisions. At this point it might not be justified to assume that
  both can be treated in the same way. Similarly to the approximation
  of the $i\mathbf{k}$ error formula, it would be useful to do
  calculations at higher precision than those presented in the paper,
  and compare them. }

As far as we concerned,
direct error estimates with the higher order terms are not easy.
Therefore, we investigate this issue in another way:
we compare the estimate (with non-bonded nearest neighbor
approximation) with the real error of the randomized COM test
(see the description of this test in page 25).
Since the only approximation made in the estimate is the leading order
approximation, the difference between the estimate and the real error
of randomized COM test is actually the contribution from the higher
order terms discarded by (48) and (53).
In Fig. 9-12,
we check this difference in a wide parameter range, and see in most
cases the estimate overlaps with the real error, which means
the leading order is a good approximation. 
However,
we do observe exceptions
at large $\beta$ when the grid spacing is $h=0.248\:\textsf{nm}$,
see Fig.11, and
at small $\beta$ when the interpolation order is $n=4$,
see Fig.12.
That means the higher order terms may play a role in the staggered
mesh interpolation errors. It also explains the unexpected discrepancy
between the staggered mesh ik- and analytical differentiation
at small $\beta$, $n=4$.
\\

% if the lead order
% is a good approximation, then two consequences are implied: (1) the
% equivalency of the staggered mesh ik- and analytical differentiation;
% (2) the overlapping of the estimate (with non-bonded nearest neighbor
% approximation) with the real error of the randomized COM testing case
% (see the description of this testing case in page 25).  In Fig. 9-12
% we check these two points for a wide parameter range, and see in most
% cases they are correct. This provide a strong evidence for the
% justification of the leading order approximation.

\textit{ The manuscript contains a few misprints, and the language
  could be polished on a few occasions by a native speaker.  }

@@

% \begin{figure}
%   \centering
%   \includegraphics[width=0.7\textwidth]{correlation.eps}
%   \caption{The schematic plot of two interacting water molecules.
%     When we calculate the force error of the left oxygen atom, the
%     correlation with its bonded hydrogens are not considered, but
%     the correlations within any neighboring molecule are considered,
%     as indicated by the solid lines with two arrows.}
%   \label{fig:1}
% \end{figure}

We define the force error by inserting a testing charge into the
considered system, and comparing the calculated and exact force on
it. In the considered water system (Example 3), the subtle difference
between this setting to the realistic case is that the testing charge
has no correlation with the charges in the system, while any real
charge is bonded to two neighboring charges (via H-O bond and H-O-H
angle), as is illustrated by the new Fig.1.
% Therefore, the error estimate does not consider the
% correlation between a charge and the its bonded neighbors, but it does
% consider all bonding correlations within any neighboring molecule (see
% new Fig.1 of the paper).
This definition is actually consistent with the usually
setting of bonded pair atoms, as mentioned by the reviewer, the
electrostatic interaction between them is not calculated.
Throughout this paper, we remove this  interaction from the
standard long-range algorithms.

% We therefore added the following sentence to make this point clearer:
% \textbf{
%   We define the force error by inserting a
% testing charge into the water system, and comparing the
% calculated and exact force on it. This definition implicitly assumes
% the independency between the testing particle and the
% system. Therefore, it sould be noted that the correlation error
% estimate (20)
% neglects the possible bonding
% of the testing particle to any hydrogen/oxygen atom, and
% calculates all bonded correlations
% within its {neighboring} molecules. This is consistent
% with the usual setting of the bonded atoms: the electrostatic interaction
% between them is considered to be \emph{not} calculated.}

@@

As far as we concerned, it is not possible to test this point
directly, but we think it is a likely explanation to the observed
phenomenon.  Consider the extreme case: within one molecule, the
hydrogen and oxygen atoms locate at the same point (strongest
correlation), then the positive and negative charges cancel, so the
error vanishes. In the case of TIP3P water model, the H-O distance is 0.1~nm,
which is much smaller than the typical
center-of-mass distance between two neighboring molecules (0.31~nm),
therefore, the bonded correlation tends to neutralize the molecule
and reduce the force error.
We have added this point to the paper.
\\
